D3064R0: How to Avoid OOTA Without Really Trying

### 2024-04-03 ootavacuous.tex
Apply Alan Stern's April Fool's feedback
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

This paper will follow P2055R0~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
in using the term \emph{full C++} to denote the standard including the
prohibition of OOTA mentioned above.
{\it Contra\/} that article, we will use the term \emph{loose C++}
(rather than \emph{strict C++}, which seems too similar to
\emph{full C++} for comfort)
to denote a hypothetical standard that excludes this prohibition but
is otherwise identical to full C++.
Unqualified \emph{C++} means full C++.

@@

For example, nonvolatile relaxed atomic accesses are subject to the
C++ memory model.
The ``order-preserving'' bullet point from
Section~\ref{sec:Volatile and Quasi Volatile Accesses}
prohibits reordering of quasi-volatile accesses when they are
to the same object,
but such reordering is in any case directly forbidden by the memory model
(the four coherence rules in 6.9.2.2p14--18 \co{[intro.races]}).
The ``onto'' bullet point also largely reiterates restrictions that follow
from the standard, in particular,
that compilers should not invent atomic stores,
duplicate atomic stores, or invent atomic loads.
On the other hand, it is permissible for compilers to omit redundant
atomic stores or fuse nonvolatile atomic accesses of adjacent objects
under appropriate circumstances;
these topics correspond to the ``total'' and ``one-to-one'' bullet points of
Section~\ref{sec:Volatile and Quasi Volatile Accesses}.

@@

Acting as if a redundant store was omitted from the source
code cannot create an OOTA cycle, because any valid execution (OOTA or not)
of the program with the first of a back-to-back pair of nonvolatile
relaxed atomic stores omitted is a valid execution of the
program with the store present.
% This is awkward because a reader might reasonably extend the "Acting
% as if" all the way to the comma.  The best I could do was "It is not
% possible to create an OOTA cycle by acting as if a redundant store
% was omitted from the source code", which is admittedly not all that
% great.

@@

This paper assumes that the user defines threads, and that the
C++ implementation executes those threads unchanged.
Future analyses might examine the possibility of ``flattening''
optimizations that combine multiple threads into one, but note
that such optimizations are more difficult than they first might
appear.

@@

Both the ARM and PowerPC architectural memory models permit reordering prior loads
against subsequent stores, but actual tests on recent hardware fail to
produce any evidence of such reordering.
Nevertheless, thus far neither ARM nor PowerPC hardware architects
have been willing to strengthen their memory models so as to forbid
such reordering, despite a number requests to do
so~\cite{RichardGrisenthwaite2023TightenRelaxedJustSayNo}.

>>>>

This paper will follow P2055R0~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
in using the term \emph{full C++} to denote the standard including the
prohibition of OOTA mentioned above.
Unlike that article, we will use the term \emph{loose C++}
(rather than \emph{strict C++}, which seems too similar to
\emph{full C++} for comfort)
to denote a hypothetical standard that excludes this prohibition but
is otherwise identical to full C++.
Unqualified \emph{C++} means full C++.

@@

For example, nonvolatile relaxed atomic accesses are subject to the
C++ memory model.
The ``order-preserving'' bullet point from
Section~\ref{sec:Volatile and Quasi Volatile Accesses}
prohibits reordering of quasi-volatile accesses when they are
to the same object,
but such reordering is in any case directly forbidden by the memory model
(the four coherence rules discussed in 6.9.2.2p14--19 \co{[intro.races]}).
The ``onto'' bullet point also largely reiterates restrictions that follow
from the standard, in particular,
that compilers should not invent atomic stores,
duplicate atomic stores, or invent atomic loads.
On the other hand, it is permissible for compilers to omit redundant
atomic stores or fuse nonvolatile atomic accesses of adjacent objects
under appropriate circumstances;
these topics correspond to the ``total'' and ``one-to-one'' bullet points of
Section~\ref{sec:Volatile and Quasi Volatile Accesses}.

@@

Omitting a redundant store cannot create an OOTA cycle,
because any valid execution (OOTA or not)
of the program with the first of a back-to-back pair of nonvolatile
relaxed atomic stores omitted is a valid execution of the
program with the store present.
% This is awkward because a reader might reasonably extend the "Acting
% as if" all the way to the comma.  The best I could do was "It is not
% possible to create an OOTA cycle by acting as if a redundant store
% was omitted from the source code", which is admittedly not all that
% great.

@@

This paper assumes that the user defines threads, and that the
C++ implementation executes those threads unchanged.
Future analyses might examine the possibility of ``flattening''
optimizations that combine multiple threads into one.
(Note that such optimizations are more difficult than they first might appear).

@@

Both the ARM and PowerPC architectural memory models permit reordering prior loads
against subsequent stores, but actual tests on recent hardware fail to
produce any evidence of such reordering.
Nevertheless, thus far neither ARM nor PowerPC hardware architects
have been willing to strengthen their memory models so as to forbid
such reordering, despite a number of requests to do
so~\cite{RichardGrisenthwaite2023TightenRelaxedJustSayNo}.
